Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-31 manuscript.tex
latest

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.

@@

We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).

@@

The model developed here is a density-dependent generalization of the classic lottery model (Results section).

Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was

Our model differs from both Grime's C/S/R and MacArthur's $r$/$K$ schemes in the role of the propagule production rate $b$, a measure of intrinsic fecundity closely related to the growth rate at low densities $r=b-d$. In both of those schemes, the essential feature of life at high densities is competition. This is less  in the $r$/$K$ scheme, which does not explicitly  it is better to have a

contributes just as much to

In the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with ``$K$-selection'' (i.e. selection at high densities) for growth ability traits.

Specifically, positive correlations between measures of $r$ and $K$ are common, both between species and strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1979,luckinbill_1978}). From the perspective of our model, this correlation is not at all surprising;

r-K correlation, meaning of K selection

>>>>

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.

@@

We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig. 4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig. \ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.

Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$. In reality there may be trade-offs between $b$ and $c$ e.g. due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}. Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

The importance of $b$ at all densities is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript. The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition. In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from
